Eco-evolutionary dynamics of finite populations from first principles

### 2023-04-19 ind_files/05_unification_discussion.tex
Minor grammatical fixes

Once again, terms of equation \eqref{nD_stochastic_Price_variance} lend themselves to straightforward biological interpretation. The quantity $(f_i-\overline{f})^2$ is a measure of the distance of $f_i$ from the population mean value $\overline{f}$, and thus covariance with $(f-\overline{f})^2$ quantifies the type of selection operating in the system: A negative correlation is indicative of stabilizing selection, and a positive correlation is indicative of disruptive (\emph{i.e.} diversifying) selection. An extreme case of diversifying selection for fitness occurs if the mean fitness is at a local minimum for fitness but $f_i \not\equiv \overline{f}$ (\emph{i.e.} the population still exhibits some variation in $f$). In this case, if the variation in $f$ is associated with a variation in fitness, then $\textrm{Cov}(w,(f - \overline{f})^2)$ is strongly positive and the population experiences a sudden explosion in variance, causing the emergence of polymorphism in the population. If $\textrm{Cov}(w,(f - \overline{f})^2)$ is still positive after the initial emergence of multiple morphs, evolution eventually leads to the emergence of stable coexisting polymorphisms in the population - This phenomenon is a slight generalization of the idea of evolutionary branching that occurs in frameworks such as adaptive dynamics \citep{geritz_evolutionarily_1998}, and a special case of the exact equation \eqref{nD_stochastic_Price_variance} has been used before to study evolutionary branching in finite populations~\citep{debarre_evolutionary_2016}.

@@

The $\overline{\tau}\sigma^2_{f}$ term quantifies the loss of variation due to stochastic extinctions (i.e. demographic stochasticity) and thus represents the classic effect of neutral genetic (or ecological) drift in finite populations. To see this, it is instructive to consider the case in which this is the only force at play. Let us imagine a population of asexual organisms in which each $f_i$ is simply a label or mark arbitrarily assigned to individuals in the population at the start of an experiment/observational study and subsequently passed on to offspring - For example, a neutral genetic tag in a part of the genome that experiences a negligible mutation rate. Let us set $\mu = 0$ so that the labels cannot change between parents to offspring. This means that we have $M_{\sigma^2_f}(\mathbf{p},N_K) \equiv 0$. Further, since the labels are arbitrary and have no effect whatsoever on the biology of the organisms, each label has the same fitness $w_i \equiv w$ and per-capita turnover $\tau_i \equiv \tau$, and thus $\overline{w} = w$ and $\overline{\tau} = \tau$. Note that since every type has the same fitness and turnover rate, we have $\textrm{Cov}\left(w,(f - \overline{f})^2\right) \equiv \textrm{Cov}\left(\tau,(f - \overline{f})^2\right) \equiv 0$. Since the labels do not change over time, we also have $\textrm{Cov}\left(\partial f/\partial t,f\right) = 0$. From \eqref{nD_stochastic_Price_variance}, we see that in this case, the variance changes as
\begin{equation}
d\sigma^2_f = - \frac{\tau\sigma^2_{f}}{KN_K}dt + \frac{1}{\sqrt{K}N_{K}(t)}dW_{\sigma^2_{f}}
\end{equation}
Where I have written $\overline{\tau} = \tau$, since every type has the same turnover rate. On taking expectations, the second term on the RHS vanishes and we see that the expected variance in the population obeys
\begin{equation}
\label{neutral_example_for_variances}
\frac{d \mathbb{E}[\sigma^2_f]}{dt} = - \left(\mathbb{E}\left[\frac{\tau}{KN_K}\right]\right)\mathbb{E}[\sigma^2_{f}]
\end{equation}
where I have decomposed the expectation of the product on the RHS into a product of expectations, which is admissible since the label $f$ is completely arbitrary and thus independent of both $\overline{\tau}$ and $N_K(t)$. Equation \eqref{neutral_example_for_variances} is a simple linear ODE and has the solution
\begin{equation}
\mathbb{E}[\sigma^2_f](t) = \sigma^2_f(0)e^{-\mathbb{E}\left[\frac{\tau}{KN_K}\right]t}
\end{equation}
which tells us that the expected diversity (variance) of labels in the population decreases exponentially over time. The rate of loss is $\mathbb{E}\left[\tau(KN_K)^{-1}\right]$, and thus, populations with higher turnover rate $\tau$  and/or lower population size $KN_K$ lose diversity faster. This is because populations with higher $\tau$ experience more stochastic events per unit time (a gambler's ruin type scenario), while extinction is `easier' in smaller populations because a smaller number of deaths is required to eliminate a label from the population completely. Note that \emph{which} labels/individuals are lost is entirely random (since all labels are arbitrary), but nevertheless, labels tend to be stochastically lost until only a small number of labels remain in the population.

@@

If the `intuitive' version of It\^{o}'s formula for $\mathcal{M}_K(\mathcal{T})$ valued stochastic processes obtained by informally `taking the limit' $m \to \infty$ in the It\^{o}'s formula for $m$-dimensional stochastic processes holds, then the exact same steps carried out in Appendices \ref{App_density_to_freq} and \ref{App_stoch_var_eqns} will `go through' essentially unchanged for quantitative traits and yield the equations obtained by simply informally taking $m \to \infty$ in \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} as the `fundamental equations' for the evolution of quantitative traits. In their supplementary sections S1 and S2,~\citet{week_white_2021} have recently proposed exactly the It\^o formula we would need for $L^2(\mathbb{R}, m)$ valued processes in which the $\mathcal{A}^{\pm}(x|\phi)$ (in my notation as defined in chapter \ref{chap_infD_processes}) are Hilbert-Schmidt operators via a heuristic It\^o multiplication table, and have shown that their heuristics are equivalent to the rigorous infinite-dimensional stochastic calculus proposed by~\citet{da_prato_stochastic_2014} for more general Hilbert space valued processes\footnote{If this sentence sounds like abstract gibberish to you, don't worry too much about the details - The essence is that~\citet{week_white_2021} have proven the `correct' formula we need holds whenever $b(x|\nu), d(x|\nu)$ and $\mathcal{M}_K(\mathcal{T})$ together fulfill certain technical requirements. My heuristics here are only a `first step' and have focused on accessibility and actually formulating the problem in formal terms rather than on mathematical propriety, and thus make no attempt to verify whether and when these technical requirements are satisfied. The required properties will likely be fulfilled only for some `nice' subspace of $\mathcal{M}_K(\mathcal{T})$ and not the entire space, and future, more rigorous work should focus on figuring out which functions are included in this subspace.}. Furthermore, using these `spacetime white noise heuristics' together with a particular functional form for (weak) solutions to an SPDE describing populations with Gaussian trait distributions,~\citet{week_white_2021} have arrived at precisely the infinite-dimensional version of equations \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} obtained by `taking the limit' $m \to \infty$ in \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} for the special case in which the type-level quantity $f$ is simply the value of the quantitative trait under study from a completely different starting point(!). Equation \eqref{disc_functional_field_eqns} is the `functional Fokker-Planck view' of the stochastic processes for the evolutionary ecology of quantitative traits that~\citet{week_white_2021} study from the `SPDE view'. Indeed, the two approaches are exactly complementary, and my formulation in chapter \ref{chap_infD_processes} provides an alternate method of attack for the study of quantitative traits in finite fluctuating populations that may be more appropriate for some particular problems, while~\posscite{week_white_2021} approach may be more appropriate for others.

@@

Currently, (stochastic) field equations are primarily used by physicists working in areas such as statistical field theory and quantum field theory, and are largely attacked using ingenious and often somewhat system-specific tools that may not necessarily be mathematically well understood~(\cite{carmona_stochastic_1999}; but see \cite{bogachev_fokker-planck-kolmogorov_2015} for a rigorous treatment of some infinite-dimensional Fokker-Planck equations). Equation \eqref{disc_functional_field_eqns} opens up the study of quantitative trait dynamics in finite fluctuating populations to analysis using some of these tools - For example, a careful reading of the literature provides many hints (albeit couched in physics language) for how (quantum) field theoretic ideas such as the Fock space representation~\citep{del_razo_probabilistic_2022}, `loop diagrams/expansions'~\citep{hochberg_effective_1999,dodd_many-body_2009}, and the often closely related path integral formalism~\citep{doi_second_1976, peliti_path_1985,hochberg_effective_1999, chow_path_2015, weber_master_2017}, can all be co-opted to coax biological insights from the processes modeled by equation \eqref{disc_functional_field_eqns}. The formalism developed in Chapter \ref{chap_infD_processes} is intended to encourage the use of such techniques and make some of the models and ideas more accessible to theorists without formal training in mathematical areas such as measure theory and functional analysis.

>>>>

Once again, terms of equation \eqref{nD_stochastic_Price_variance} lend themselves to straightforward biological interpretation. The quantity $(f_i-\overline{f})^2$ is a measure of the distance of $f_i$ from the population mean value $\overline{f}$, and thus covariance with $(f-\overline{f})^2$ quantifies the type of selection operating in the system: A negative correlation is indicative of stabilizing selection, and a positive correlation is indicative of disruptive (\emph{i.e.} diversifying) selection. An extreme case of diversifying selection for fitness occurs if the mean fitness is at a local minimum for fitness but $f_i \not\equiv \overline{f}$ (\emph{i.e.} the population still exhibits some variation in $f$). In this case, if the variation in $f$ is associated with a variation in fitness, then $\textrm{Cov}(w,(f - \overline{f})^2)$ is strongly positive and the population experiences a sudden explosion in variance, causing the emergence of polymorphism in the population. If $\textrm{Cov}(w,(f - \overline{f})^2)$ is still positive after the initial emergence of multiple morphs, evolution eventually leads to the emergence of stable coexisting polymorphisms in the population - this phenomenon is a slight generalization of the idea of evolutionary branching that occurs in frameworks such as adaptive dynamics \citep{geritz_evolutionarily_1998}, and a special case of the exact equation \eqref{nD_stochastic_Price_variance} has been used before to study evolutionary branching in finite populations~\citep{debarre_evolutionary_2016}.

@@

The $\overline{\tau}\sigma^2_{f}$ term quantifies the loss of variation due to stochastic extinctions (i.e. demographic stochasticity) and thus represents the classic effect of neutral genetic (or ecological) drift in finite populations. To see this, it is instructive to consider the case in which this is the only force at play. Let us imagine a population of asexual organisms in which each $f_i$ is simply a label or mark arbitrarily assigned to individuals in the population at the start of an experiment/observational study and subsequently passed on to offspring - for example, a neutral genetic tag in a part of the genome that experiences a negligible mutation rate. Let us set $\mu = 0$ so that the labels cannot change between parents to offspring. This means that we have $M_{\sigma^2_f}(\mathbf{p},N_K) \equiv 0$. Further, since the labels are arbitrary and have no effect whatsoever on the biology of the organisms, each label has the same fitness $w_i \equiv w$ and per-capita turnover $\tau_i \equiv \tau$, and thus $\overline{w} = w$ and $\overline{\tau} = \tau$. Note that since every type has the same fitness and turnover rate, we have $\textrm{Cov}\left(w,(f - \overline{f})^2\right) \equiv \textrm{Cov}\left(\tau,(f - \overline{f})^2\right) \equiv 0$. Since the labels do not change over time, we also have $\textrm{Cov}\left(\partial f/\partial t,f\right) = 0$. From \eqref{nD_stochastic_Price_variance}, we see that in this case, the variance changes as
\begin{equation}
d\sigma^2_f = - \frac{\tau\sigma^2_{f}}{KN_K}dt + \frac{1}{\sqrt{K}N_{K}(t)}dW_{\sigma^2_{f}}
\end{equation}
Where I have written $\overline{\tau} = \tau$, since every type has the same turnover rate. On taking expectations, the second term on the RHS vanishes and we see that the expected variance in the population obeys
\begin{equation}
\label{neutral_example_for_variances}
\frac{d \mathbb{E}[\sigma^2_f]}{dt} = - \left(\mathbb{E}\left[\frac{\tau}{KN_K}\right]\right)\mathbb{E}[\sigma^2_{f}]
\end{equation}
where I have decomposed the expectation of the product on the RHS into a product of expectations, which is admissible since the label $f$ is completely arbitrary and thus independent of both $\overline{\tau}$ and $N_K(t)$. Equation \eqref{neutral_example_for_variances} is a simple linear ODE and has the solution
\begin{equation}
\mathbb{E}[\sigma^2_f](t) = \sigma^2_f(0)e^{-\mathbb{E}\left[\frac{\tau}{KN_K}\right]t}
\end{equation}
which tells us that the expected diversity (variance) of labels in the population decreases exponentially over time. The rate of loss is $\mathbb{E}\left[\tau(KN_K)^{-1}\right]$, and thus, populations with higher turnover rate $\tau$  and/or lower population size $KN_K$ lose diversity faster. This is because populations with higher $\tau$ experience more stochastic events per unit time (a gambler's ruin type scenario), while extinction is `easier' in smaller populations because a smaller number of deaths is required to eliminate a label from the population completely. Note that \emph{which} labels/individuals are lost is entirely random (since all labels are arbitrary), but nevertheless, labels tend to be stochastically lost until only a small number of labels remain in the population.

@@

If the `intuitive' version of It\^{o}'s formula for $\mathcal{M}_K(\mathcal{T})$ valued stochastic processes obtained by informally `taking the limit' $m \to \infty$ in the It\^{o}'s formula for $m$-dimensional stochastic processes holds, then the exact same steps carried out in Appendices \ref{App_density_to_freq} and \ref{App_stoch_var_eqns} will `go through' essentially unchanged for quantitative traits and yield the equations obtained by simply informally taking $m \to \infty$ in \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} as the `fundamental equations' for the evolution of quantitative traits. In their supplementary sections S1 and S2,~\citet{week_white_2021} have recently proposed exactly the It\^o formula we would need for $L^2(\mathbb{R}, m)$ valued processes in which the $\mathcal{A}^{\pm}(x|\phi)$ (in my notation as defined in chapter \ref{chap_infD_processes}) are Hilbert-Schmidt operators via a heuristic It\^o multiplication table, and have shown that their heuristics are equivalent to the rigorous infinite-dimensional stochastic calculus proposed by~\citet{da_prato_stochastic_2014} for more general Hilbert space valued processes\footnote{If this sentence sounds like abstract gibberish to you, don't worry too much about the details -the essence is that~\citet{week_white_2021} have proven the `correct' formula we need holds whenever $b(x|\nu), d(x|\nu)$ and $\mathcal{M}_K(\mathcal{T})$ together fulfill certain technical requirements. My heuristics here are only a `first step' and have focused on accessibility and actually formulating the problem in formal terms rather than on mathematical propriety, and thus make no attempt to verify whether and when these technical requirements are satisfied. The required properties will likely be fulfilled only for some `nice' subspace of $\mathcal{M}_K(\mathcal{T})$ and not the entire space, and future, more rigorous work should focus on figuring out which functions are included in this subspace.}. Furthermore, using these `spacetime white noise heuristics' together with a particular functional form for (weak) solutions to an SPDE describing populations with Gaussian trait distributions,~\citet{week_white_2021} have arrived at precisely the infinite-dimensional version of equations \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} obtained by `taking the limit' $m \to \infty$ in \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} for the special case in which the type-level quantity $f$ is simply the value of the quantitative trait under study from a completely different starting point(!). Equation \eqref{disc_functional_field_eqns} is the `functional Fokker-Planck view' of the stochastic processes for the evolutionary ecology of quantitative traits that~\citet{week_white_2021} study from the `SPDE view'. Indeed, the two approaches are exactly complementary, and my formulation in chapter \ref{chap_infD_processes} provides an alternate method of attack for the study of quantitative traits in finite fluctuating populations that may be more appropriate for some particular problems, while~\posscite{week_white_2021} approach may be more appropriate for others.

@@

Currently, (stochastic) field equations are primarily used by physicists working in areas such as statistical field theory and quantum field theory, and are largely attacked using ingenious and often somewhat system-specific tools that may not necessarily be mathematically well understood~(\cite{carmona_stochastic_1999}; but see \cite{bogachev_fokker-planck-kolmogorov_2015} for a rigorous treatment of some infinite-dimensional Fokker-Planck equations). Equation \eqref{disc_functional_field_eqns} opens up the study of quantitative trait dynamics in finite fluctuating populations to analysis using some of these tools - for example, a careful reading of the literature provides many hints (albeit couched in physics language) for how (quantum) field theoretic ideas such as the Fock space representation~\citep{del_razo_probabilistic_2022}, `loop diagrams/expansions'~\citep{hochberg_effective_1999,dodd_many-body_2009}, and the often closely related path integral formalism~\citep{doi_second_1976, peliti_path_1985,hochberg_effective_1999, chow_path_2015, weber_master_2017}, can all be co-opted to coax biological insights from the processes modeled by equation \eqref{disc_functional_field_eqns}. The formalism developed in Chapter \ref{chap_infD_processes} is intended to encourage the use of such techniques and make some of the models and ideas more accessible to theorists without formal training in mathematical areas such as measure theory and functional analysis.

### 2023-04-19 ind_files/06_nontechnical_discussion.tex
Minor grammatical fixes

In this thesis, I present a mathematical framework for such a reformulation. Part \ref{part_theory} of this thesis develops a formalism for both populations of individuals that vary in discrete characters (Chapter \ref{chap_BD}) as well as populations of individuals that vary in a single one-dimensional quantitative character (Chapter \ref{chap_infD_processes}). The central result of this reformulation is a series of stochastic differential equations derived in Chapter \ref{chap_unification}. To begin, I derive an equation for change in type frequencies in the population (equation \eqref{nD_stochastic_RM}) that generalizes the replicator-mutator equation to finite, fluctuating, closed populations evolving in continuous time. From this, I show how one can derive an equation for changes in the population mean of an arbitrary type-level quantity (equation \eqref{nD_stochastic_Price}) that generalizes the dynamic, continuous time version of the Price equation to finite, fluctuating populations. I also derive an equation for changes in the population variance of an arbitrary type-level quantity (equation \eqref{nD_stochastic_Price_variance}) in such populations that generalizes a recent infinite population formulation of the same~\citep{lion_theoretical_2018}. My work thus generalizes some fundamental formal structures of eco-evolutionary population dynamics to finite, fluctuating populations. In particular, my equations generalize the unifying formalism described in~\cite{lion_theoretical_2018}\footnote{which itself is a reformulation of the relatively well-known unification of eco-evolutionary dynamics via the Price equation~\citep{frank_natural_2012, queller_fundamental_2017, luque_mirror_2021} in a dynamically sufficient, continuous time framework} to finite, fluctuating populations - Taking $K \to \infty$ in equations \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price}, and \eqref{nD_stochastic_Price_variance} recover equations (6), (11), and (14) in~\cite{lion_theoretical_2018} respectively as their infinite population limits. While my equations thus recover standard results such as the Price equation and the replicator-mutator equation in the infinite population limit, they also predict that these results do not completely capture the behavior of finite populations.

@@

Lastly, noise-induced selection is particular to fluctuating populations and does not occur in models with fixed population sizes\footnote{If $\sum_j x_j$ is a constant, the map $x_i \to x_i/\sum_j x_j$ becomes a linear map and we no longer need It\^o's formula to move from densities to frequencies in the derivation I conduct in Appendix \ref{App_density_to_freq}; Thus, simply dividing equation \eqref{nD_Ito_SDE} by the (now constant) total population size provides the complete dynamics of the system in frequency space: Note that the directional terms in equation \eqref{nD_Ito_SDE} depend only on $\mathbf{A}^-$, which in turn depends only on the fitness $w_i$ and the mutation terms, and this system thus has no noise-induced selection.} such as Wright-Fisher or Moran models, suggesting that working with such constant population frameworks is not sufficient to accurately capture the dynamics of real populations. Several theoretical studies have pointed out that the empirically common practice of approximating the size of fluctuating populations through a constant `effective population size' (often obtained as the harmonic mean of population size over time) is of limited applicability~\citep{gillespie_natural_1974,sjodin_meaning_2005,parsons_consequences_2010,iizuka_effective_2010,abu_awad_effects_2018,kuosmanen_turnover_2022}. Recent experimental evolution studies have also directly shown that the harmonic mean of population size need not be a good proxy for capturing and/or predicting evolutionary dynamics in fluctuating populations~\citep{chavhan_larger_2019}. In our case, noise-induced selection cannot be captured by any constant effective population size, harmonic mean or otherwise, since noise-induced selection effects also depend on the turnover $\tau_i$ of each type and are thus a property of each \emph{type} in a fluctuating population (and not merely a function of the population size as a whole), meaning that each type would demand its own effective population size for the entire population. My results thus lend further support to one of the general messages of these previous studies - Approximating fluctuating populations via a simpler constant (effective) population size requires some careful justification, since it may inadvertently remove important evolutionary properties of the systems under study.

@@

~\citet{lion_theoretical_2018} has pointed out that in the dynamic setting (for infinite populations), the replicator-mutator equation \eqref{nD_replicator_mutator} is in some sense the `most fundamental' equation, and equations like the dynamically sufficient version of the Price equation are best viewed as a hierarchy of moment equations for the population mean, population variance, etc. of a type-level quantity. This is also true in my framework - Equation \eqref{nD_stochastic_RM} is the most fundamental equation for population dynamics, and equations like \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} can then be derived from \eqref{nD_stochastic_RM} through repeated application of It\^o's formula, in principle for any moment of the distribution of the quantity $f$ in the population (though this quickly becomes too tedious to actually carry out in practice). If we additionally assume that the quantity $f$ follows a Gaussian distribution in the population, then the mean and variance completely characterize the distribution of $f$, and equations \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price}, and \eqref{nD_stochastic_Price_variance} together specify the complete stochastic dynamics of the system. Note that even in this case, actually solving these equations analytically for equilibrium/stationary state distributions of $\mathbf{p}$, $\overline{f}$, and $\sigma^2_f$ will quickly become impossibly difficult if the birth and death rate functions are complicated. Indeed, previous studies indicate that in high dimensions, evolutionary birth-death models can exhibit a dizzying array of complicated behavior, including limit cycles and chaotic trajectories~\citep{doebeli_diversity_2017}. However, while this may present impediments to formulating exact solutions to the equations I formulate, the point of these equations is not necessarily to solve them to begin with.

@@

My formalism here is only a first step, and there are many biologically important factors that I have neglected in this thesis. My framework works with unstructured populations, neglecting any potential effects of groups, age, class, sex, developmental stage, or space\footnote{Of course,  since position in continuous space is just a special case of a quantitative trait, we have technically incorporated space in a very limited sense: The formalism of chapter \ref{chap_infD_processes} can equally well describe a finite population of exactly identical individuals moving through one-dimensional space, and so can technically describe phenomena like range expansion of clonal populations in one dimension. However, I am only noting this as a technicality - Such populations are likely somewhat `evolutionarily boring', since all individuals in the model must always be exactly identical in all aspects other than spatial location.}, all of which can lead to very complex and often surprising dynamics. Explicitly incorporating features such as population structure, sex, and space from first principles in an analytically tractable framework is a formidable task that may need innovative new mathematical and biological arguments, and presents a fantastic opportunity for future studies. I have also neglected any potential complications introduced by genotype-phenotype maps and genetic processes such as dominance, epistasis, and pleiotropy since my assumptions on the functional forms of the birth and death rates (\eqref{nD_functional_forms_for_replicator} or \eqref{BD_for_kimura}) should be general enough to incorporate these effects (in principle) while analyzing specific models.

>>>>

In this thesis, I present a mathematical framework for such a reformulation. Part \ref{part_theory} of this thesis develops a formalism for both populations of individuals that vary in discrete characters (Chapter \ref{chap_BD}) as well as populations of individuals that vary in a single one-dimensional quantitative character (Chapter \ref{chap_infD_processes}). The central result of this reformulation is a series of stochastic differential equations derived in Chapter \ref{chap_unification}. To begin, I derive an equation for change in type frequencies in the population (equation \eqref{nD_stochastic_RM}) that generalizes the replicator-mutator equation to finite, fluctuating, closed populations evolving in continuous time. From this, I show how one can derive an equation for changes in the population mean of an arbitrary type-level quantity (equation \eqref{nD_stochastic_Price}) that generalizes the dynamic, continuous time version of the Price equation to finite, fluctuating populations. I also derive an equation for changes in the population variance of an arbitrary type-level quantity (equation \eqref{nD_stochastic_Price_variance}) in such populations that generalizes a recent infinite population formulation of the same~\citep{lion_theoretical_2018}. My work thus generalizes some fundamental formal structures of eco-evolutionary population dynamics to finite, fluctuating populations. In particular, my equations generalize the unifying formalism described in~\cite{lion_theoretical_2018}\footnote{which itself is a reformulation of the relatively well-known unification of eco-evolutionary dynamics via the Price equation~\citep{frank_natural_2012, queller_fundamental_2017, luque_mirror_2021} in a dynamically sufficient, continuous time framework} to finite, fluctuating populations - taking $K \to \infty$ in equations \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price}, and \eqref{nD_stochastic_Price_variance} recover equations (6), (11), and (14) in~\cite{lion_theoretical_2018} respectively as their infinite population limits. While my equations thus recover standard results such as the Price equation and the replicator-mutator equation in the infinite population limit, they also predict that these results do not completely capture the behavior of finite populations.

@@

Lastly, noise-induced selection is particular to fluctuating populations and does not occur in models with fixed population sizes\footnote{If $\sum_j x_j$ is a constant, the map $x_i \to x_i/\sum_j x_j$ becomes a linear map and we no longer need It\^o's formula to move from densities to frequencies in the derivation I conduct in Appendix \ref{App_density_to_freq}; Thus, simply dividing equation \eqref{nD_Ito_SDE} by the (now constant) total population size provides the complete dynamics of the system in frequency space: Note that the directional terms in equation \eqref{nD_Ito_SDE} depend only on $\mathbf{A}^-$, which in turn depends only on the fitness $w_i$ and the mutation terms, and this system thus has no noise-induced selection.} such as Wright-Fisher or Moran models, suggesting that working with such constant population frameworks is not sufficient to accurately capture the dynamics of real populations. Several theoretical studies have pointed out that the empirically common practice of approximating the size of fluctuating populations through a constant `effective population size' (often obtained as the harmonic mean of population size over time) is of limited applicability~\citep{gillespie_natural_1974,sjodin_meaning_2005,parsons_consequences_2010,iizuka_effective_2010,abu_awad_effects_2018,kuosmanen_turnover_2022}. Recent experimental evolution studies have also directly shown that the harmonic mean of population size need not be a good proxy for capturing and/or predicting evolutionary dynamics in fluctuating populations~\citep{chavhan_larger_2019}. In our case, noise-induced selection cannot be captured by any constant effective population size, harmonic mean or otherwise, since noise-induced selection effects also depend on the turnover $\tau_i$ of each type and are thus a property of each \emph{type} in a fluctuating population (and not merely a function of the population size as a whole), meaning that each type would demand its own effective population size for the entire population. My results thus lend further support to one of the general messages of these previous studies - approximating fluctuating populations via a simpler constant (effective) population size requires some careful justification, since it may inadvertently remove important evolutionary properties of the systems under study.

@@

~\citet{lion_theoretical_2018} has pointed out that in the dynamic setting (for infinite populations), the replicator-mutator equation \eqref{nD_replicator_mutator} is in some sense the `most fundamental' equation, and equations like the dynamically sufficient version of the Price equation are best viewed as a hierarchy of moment equations for the population mean, population variance, etc. of a type-level quantity. This is also true in my framework - equation \eqref{nD_stochastic_RM} is the most fundamental equation for population dynamics, and equations like \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} can then be derived from \eqref{nD_stochastic_RM} through repeated application of It\^o's formula, in principle for any moment of the distribution of the quantity $f$ in the population (though this quickly becomes too tedious to actually carry out in practice). If we additionally assume that the quantity $f$ follows a Gaussian distribution in the population, then the mean and variance completely characterize the distribution of $f$, and equations \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price}, and \eqref{nD_stochastic_Price_variance} together specify the complete stochastic dynamics of the system. Note that even in this case, actually solving these equations analytically for equilibrium/stationary state distributions of $\mathbf{p}$, $\overline{f}$, and $\sigma^2_f$ will quickly become impossibly difficult if the birth and death rate functions are complicated. Indeed, previous studies indicate that in high dimensions, evolutionary birth-death models can exhibit a dizzying array of complicated behavior, including limit cycles and chaotic trajectories~\citep{doebeli_diversity_2017}. However, while this may present impediments to formulating exact solutions to the equations I formulate, the point of these equations is not necessarily to solve them to begin with.

@@

My formalism here is only a first step, and there are many biologically important factors that I have neglected in this thesis. My framework works with unstructured populations, neglecting any potential effects of groups, age, class, sex, developmental stage, or space\footnote{Of course,  since position in continuous space is just a special case of a quantitative trait, we have technically incorporated space in a very limited sense: The formalism of chapter \ref{chap_infD_processes} can equally well describe a finite population of exactly identical individuals moving through one-dimensional space, and so can technically describe phenomena like range expansion of clonal populations in one dimension. However, I am only noting this as a technicality - such populations are likely somewhat `evolutionarily boring', since all individuals in the model must always be exactly identical in all aspects other than spatial location.}, all of which can lead to very complex and often surprising dynamics. Explicitly incorporating features such as population structure, sex, and space from first principles in an analytically tractable framework is a formidable task that may need innovative new mathematical and biological arguments, and presents a fantastic opportunity for future studies. I have also neglected any potential complications introduced by genotype-phenotype maps and genetic processes such as dominance, epistasis, and pleiotropy since my assumptions on the functional forms of the birth and death rates (\eqref{nD_functional_forms_for_replicator} or \eqref{BD_for_kimura}) should be general enough to incorporate these effects (in principle) while analyzing specific models.
